Lowestcase and Uppestcase letters: Further adventures in Derp Learning

### 2021-03-10 trunk/lowercase/paper/lowercase.tex
Some work on paper

% XXX want some macro for referring to a letter's shape, maybe
% with a box around it?

@@

\newcommand\weirdcharlo{\includegraphics[width=0.8em]{weirdchar-lo}}
\newcommand\weirdcharup{\includegraphics[width=0.8em]{weirdchar-up}}

@@

\begin{figure}[ht]
% \includegraphics[width=0.9 \linewidth]{manual}
\caption{TODO cotraining} \label{fig:cotraining}
\end{figure}

Because \makelowercase\ is getting training examples of
uppercase/lowercase pairs from real fonts, it remains grounded on real
letters. It is also free to generate new shapes in the open domain
(outside \letterform{A}--\letterform{Z}). However, it is penalized if
its behavior is not the inverse of whatever \makeuppercase\ is
currently doing. And since we do the symmetric thing for
\makeuppercase\, there is a (slow) feedback loop between the two
models that keeps them from straying too far from the grounded
examples. The idea is that this allows them to do some creative
generalization outside their native domains, but in a way that
still has some constraint.

In practice, we don't feed arbitrary shapes to the models. We
just need something letter-like, and in fact we have a large
collection of letter-like shapes among our existing fonts. So
we pass already-lowercase shapes to \makelowercase, in order
to generate inversion examples for training \makeuppercase.
These shapes are clearly letter-like (they {\em are} letters) and
are also of interest to us anyway, since we want to try to
generate lowerercase and upperercase letters with these models.

% $\langle\,$\letterform{Q}, \letterform{q}$\,\rangle$
% $\langle\,$\letterform{e}, \letterform{E}$\,\rangle$

@@

@@

% (lowercase i intelligence: redi)

% The care and feeding of sparse matrices

>>>>

% XXX introduce the meaning of the \letterform syntax somewhere?

@@

\newcommand\weirdcharlo{\includegraphics[width=0.75em]{weirdchar-lo}}
\newcommand\weirdcharup{\includegraphics[width=0.75em]{weirdchar-up}}
\newcommand\lowerlowera{\includegraphics[width=0.6em]{lowerlowera}}

@@

\begin{figure}[ht]
\includegraphics[width=0.9 \linewidth]{training}
\caption{ Simultaneously training the two models. This example
  illustrates how a pair of letterforms \letterform{A} and
  \letterform{a} from the same font becomes four training examples.
  The pair straightforwardly generates an example
  \trainingexample{A}{a} for the \makelowercase\ queue, and an example
  \trainingexample{a}{A} for the \makeuppercase\ queue. Separately, we
  supply \letterform{a} to the \makelowercase\ model, simply to get
  the current output \letterform{\lowerlowera} (no model updates are
  performed). But this pair reversed becomes a training example
  \trainingexample{\lowerlowera}{a} for the \makeuppercase\ queue.
} \label{fig:cotraining}
\end{figure}

Because \makelowercase\ is getting training examples of
uppercase/lowercase pairs from real fonts, it remains grounded on real
letters. It is also free to generate new shapes for the open domain
(outside \letterform{A}--\letterform{Z}). However, it is penalized if
its behavior is not the inverse of whatever \makeuppercase\ is
currently doing. And since we do the symmetric thing for
\makeuppercase\, there is a (slow) feedback loop between the two
models that keeps them from straying too far from the grounded
examples. The idea is that this allows them to do some creative
generalization outside their native domains, but in a way that
still has some constraint.

In practice, we don't feed arbitrary shapes to the models. We just
need something letter-like, and in fact we have a large collection of
letter-like shapes among our existing fonts! We pass already-lowercase
shapes to \makelowercase, in order to generate inversion examples for
training \makeuppercase. These shapes are clearly letter-like (they
{\em are} letters) and are also of interest to us anyway, since we
want to try to generate lowerercase and upperercase letters from
the trained models.

@@

Sprechen of fonts, I downloaded every font I could find on the whole
internet. This was overkill. The resulting directory tree contained
over 100,000 files, many of which were duplicates. Exact duplicates
are easy to find, but since many of these files were the result of 30
years of community transmission, they had acquired various mutations.
One of the first things I did was write software to automatically
remove files that were essentially duplicates even if they weren't
exactly the same bytes.

Next, my lord, do people have bad taste! And I say this as someone who
made dozens of amateurish fonts\cite{dbzfonts} as a high school and
college student and who is contributing several new questionable
fonts\label{sec:newfonts} as a result of this paper. The database is
just filled with garbage that is unusable for this project: Fonts that
are completely illegible, fonts that are missing most of their
characters, fonts with millions of control points, Comic Sans MS,
fonts where every glyph is a drawing of a train, fonts where
everything is fine except that just the lowercase r has a width of
{\tt MAX\_INT}, and so on. So I built a UI
(Figure~\ref{fig:sortition}) for efficiently and mind-numbingly
cleaning up the database by marking fonts as broken or suitable (and
also categorizing them as serif, sans-serif, decorative, techno, etc.,
which I never used). In doing this I noticed another extremely common
problem, which was that many fonts had the same letter shapes for
uppercase and lowercase letters. This would not do for the current
application!

\begin{figure*}[ht]
\includegraphics[width=0.9 \linewidth]{sortition}
\caption{ The interactive font data-cleaning UI. A seemingly endless
  series of fonts presents, with single keypresses putting the fonts
  into common categories such as (b)roken.
  % XXX maybe write more?
} \label{fig:sortition}
\end{figure*}

But why manually mark fonts with nearly the same upper- and
lowercase letters, when you could build a complicated automatic
solution? The first pass identified fonts whose letters were
exactly the same, but this was only a small fraction of the
problematic fonts. A common issue was that the lowercase characters
were very slightly modified versions of the uppercase ones, often
scaled and translated and then perhaps ``optimized'' during the
font export.

So, for a given font, I want to reject it if for most pairs of cased
letters \letterform{A},\letterform{a}, \letterform{a} is close to a
linear transformation of \letterform{A}. This problem can probably be
solved with math, but it didn't sound that fun. Instead I tried out
a new tool, and it worked well enough that I've now added it to the
permanent rotation: Black-box function optimizers.

\paragraph{Black-box optimization.} If you have a function and want
to find arguments that minimize its output, the most efficient
techniques are generally those like gradient descent. (In fact, the
backpropagation algorithm we use to training the neural network in
Section~\ref{sec:neural} is gradient descent on the function that
takes the model weights and produces an error value for each output
node.) The problem with this is that you need to do some math to
compute the gradient function, and anyway you need to deal with fiddly
bits (Section~\ref{sec:fiddly}) unless the function is convex and
smooth, which it will not be. If you don't want to deal with that,
and have a fast computer (and who doesn't?), black-box optimization
algorithms are worth considering. Here, the interface\footnote{
  Here a simplfied wrapper around BiteOpt~\cite{biteopt} in my {\tt cc-lib}
  library. See \url{https://sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/cc-lib/opt/}.}
  is just something like (C++):

\begin{verbatim}
  double Minimize1D(
      const std::function<double(double)> &f,
      double lower_bound,
      double upper_bound,
      int iters);
\end{verbatim}

which takes a function \verb+f+ of type \verb+double+ $\rightarrow$
\verb+double+, finite bounds on the argument's value, the maximum
number of times to call that function, and returns the argument it
found that produced the minimal value. Not as fast as gradient
descent, but in practice ``if the function is kinda smooth'' these
optimizers produce excellent results! The chief selling point for me
is that I don't need to think about anything except for writing the
function that I want minimized, which I just express in normal code.

\begin{figure}[ht]
\includegraphics[width=0.9 \linewidth]{casealignment}
\caption{ Example alignment to reject the font
  {\tt DrippingGooExtended}.
  At left, \letterform{A} (red) and
  \letterform{a} (green) rendered with the identity transform, and
  their alignment ($35\%$ difference) below. At right, the transform
  found by the black-box optimizer and the resulting alignment with
  $1.7\%$ difference. Note that the shapes are still not an exact
  match (the export process has to round the data to integers and
  might apply other non-linear transformations like curve
  simplification), but these are clearly not a useful pair for
  the current problem.} \label{fig:casealignment}
\end{figure}

In this case, I render the letterform \letterform{A} and then optimize
a four argument function taking {\tt xoff}, {\tt yoff}, {\tt xscale},
{\tt yscale}. This function renders \letterform{a} with those
parameters, then just computes the difference in the two rendered
bitmaps. This finds the best alignment of the two letterforms (under
the linear transformation) in a few hundred milliseconds
(Figure~\ref{fig:casealignment}). If the disagreement is low as a
function of the total pixels, then we say that the letters have the
same case. If enough of them have the same case, we reject the font. I
set the thresholds by looking at the P/R curve computed on random
hand-labeled examples (Figure~\ref{fig:samecasepr}).

\begin{figure}[ht]
\includegraphics[width=0.9 \linewidth]{samecasepr}
\caption{ Precision--recall curve for automatically detecting fonts
  that have basically the same upper- and lowercase shapes. It's good!
  This is how you want 'em to look!
} \label{fig:samecasepr}
\end{figure}

\medskip

I labeled fonts using the UI until I had 10,000 that were clean enough
for a training set and passed the same-case heuristic.

\section{The simplest thing that might work}

Before getting fancy (which we we will) it's good engineering hygiene
to try the simpliest thing that might just work (it doesn't). Fonts
are represented as vector data (lines and quadratic B\'ezier curves).
Can we just train a network that takes these lines and curves as input
and predicts the lower- or uppercased letter in the same format? (No.)

We'll at least put the data in a somewhat normalized form. The neural
network will take a fixed number of inputs to a fixed number of
outputs, so a simple approach is to decide on some maximum number
control points per letter, and only try training on fonts whose
letterforms fit within that budget. Letterforms can be made of
multiple contours (e.g. a stacked \letterform{g} typically has two
holes in it, and \letterform{j} has two disjoint parts). I found that
most clean fonts had three or fewer contours, and when sorting them by
descending length, most fit within 38, 14, and 10 endpoints for the
three. So, I only train on fonts where all of the letters fit within
this budget.\footnote{It would not be a good idea to reject only the letters
that don't fit, because it might result in the network being trained
on more \letterform{l}s (tends to be simple) than \letterform{g}s
(tends to be complex).}

% TODO: Illustrate a letterform and its contours?

Rather than try to work with both lines and B\'ezier curves, I
normalize each contour to only contain B\'eziers, by turning a line
segment into an equivalent B\'ezier with its control point at the
midpoint. This frees us from having to distinguish the two types in
the data. We also need each of the three contours to not be {\em too
  short}, so I fill out the fixed-size buffers by repeating the last
point. This is not great but does have the correct meaning (bunch of
useless zero-length edges). It has the property that any predicted
data can be rendered and has a straightforward point-by-point error
function (which might not be the case if we were predicting a dynamic
number of points). Finally, I output the contour's points starting
from the point closest to the origin in order to reduce one needless
degree of freedom.

% XXX total number of parameters?
The network I trained has an input layer size of $154 = (38 + 14 + 10)
\times 4 + 3 \times 2$ (one control point and one end point per
B\'ezier curve), plus a starting point for each of the three contours.
The output layer is the same size, plus 26 (see below). There are
three hidden layers of size 308, 308, 360. The first two layers are
dense and the third sparse (about half of the weights are zero). All
layers are leaky rectified linear units. If you're taking notes,
don't, as again this does not work well, and I don't know how people
figure out what the right network size is anyway. I just made it up.
You can give {\em me} your notes.

\newcommand\nan{\textsf{NaN}}
\renewcommand\inf{\textsf{inf}}

I trained the network using a home-grown (why??) GPU-based package
that I wrote for {\em Red i removal with artificial retina
  networks}~\cite{murphy2015redi}---an example of ``lowercase i
artificial intelligence''---and have improved as I repurposed it for
other projects, such as {\em Color- and piece-blind
  chess}~\cite{murphy2019blind}. It is ``tried and true'' in the
sense that ``every time I tried using it, I truly wanted to throw
my computer out the window, and retire to a hermitage in the glade
whenceforth I shall nevermore be haunted by a model which has
overnight become a sea of \inf{}s and \nan{}s.''

% TODO: Screenshot of this thing's output.
% TODO: Font file generated from this.
% TODO: Fitting the prediction to the actual in order to coax
% better results, failing.

% the sdf problem

\section{The care and feeding of sparse matrices} \label{sec:neural}

\label{sec:fiddly}

@@

% stretch the loss function near the onedge value
